Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-08-12 analysis/paper/paper.Rmd
CONTINUED UPDATES TO PAPER/FIGURES

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring
(April-June) as sea ice begins to melt and retreat northward. Rapid acceleration of
climate change in Arctic ecosystems is therefore of concern as the quantity and quality of
suitable habitat is forecast to decrease. In this paper, we use data collected from
satellite-linked bio-loggers deployed between 2005 and 2020 to investigate the seasonal
timing and environmental factors affecting sea ice use by seals (specifically, bearded,
ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to providing
baseline data on phenology, these data also allow us to quantify ‘availability’, which is
needed to accurately estimate abundance from aerial survey counts of seals basking on ice
(i.e., to correct for the proportion of animals that are in the water while surveys are
conducted). Using generalized linear mixed pseudo-models to properly account for temporal
autocorrelation, we fit models with a variety of covariates (e.g., day-of-year, solar
hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to
examine their ability to explain variation in hourly haul-out records. We found evidence
for strong diurnal and within-season patterns in haul-out behavior, as well as strong
weather effects (particularly wind and temperature). In general, seals were more likely to
rest on ice in the middle of the day and when wind speed was low and temperatures were
higher. Haul-out propensity increased through March and April, peaking in May and early
June before declining again. The timing and frequency of haul-out events also varied based
on species and age-sex class. For ribbon and spotted seals, models with year effects were
highly supported, indicating that the timing and magnitude of haul-out behavior varied
among years. However, we did not find evidence that haul-out timing was linked to annual
sea ice extent. Our analysis emphasizes the importance of accounting for variation in
haul-out behavior when interpreting the number of seals counted in aerial surveys.

@@

The global climate disruption is causing considerable reduction in Arctic sea ice extent,
volume, and seasonal presence [@kwok2018erl, @meier2014rg, @wang2017dsrpitsio]. These
changes have ripple effects on Arctic organisms, ecosystems, and the human communities who
live in the region [@huntington2020ncc]. Such disruptions are a particular cause of
concern for ice-associated seals (ringed (*Phoca hispida*), bearded (*Erignathus
barbatus*), spotted (*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that
depend on spring and early summer sea ice (March-June) in the Bering and Chukchi seas as a
platform for important functions, such as pupping, nursing, breeding, and molting
[@boveng2009a;
@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps complicate
predictions about the ultimate effects of changes in sea ice on the behavior, abundance,
and distribution of these seals. Knowledge about evolutionary constraints on the timing of
reproductive and molting behavior is generally lacking, so it is difficult to predict the
readiness with which ice-associated seal species will be able to adapt to future changes
(e.g., by adjusting pupping or molting schedules to earlier dates or different locales).
Additionally, trends in abundance of these species are largely unknown, so it is difficult
to say what effect, if any, declines in sea ice habitat have had, or will have, on seal
densities.

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals rest
on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The remoteness of the Bering and Chukchi
seas mean direct scientific observation of seal behavior is impractical. Bio-logging
devices are, thus, especially useful tools for collecting key information on movement and
haul-out behavior for these species. Bio-logging records of time spent out of the water
provide valuable data for identifying covariates that explain variation in haul-out
behavior. For instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied
on bio-logging data to demonstrate greater haul-out propensity in juvenile crabeater seals
(*Lobodon carcinophaga*) than adults, with highest probabilities in February and at times
close to solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated
seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson
et al. [-@bengtson2005pb] documented a higher propensity for ringed seal basking near
solar noon, as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded , ribbon ,
and spotted seals using much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that
the proportion of time bearded seals spent hauled out progressively increased through
spring and summer. And, Ver Hoef et al. [-@verhoef2014sm] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded, ribbon, and
spotted seals. However, their primary focus was to develop a correction factor for
estimating seal abundance, and like a subsequent analysis by Conn et al. [-@conn2014mee],
they did not attempt to separately analyze haul-out schedules for different age and sex
classes (although initial models with sex and age as fixed effects were insignificant at p
= 0.05 [@verhoef2014sm]).

Several authors have used logistic regression-style analyses to estimate the probability
of sea ice use from data garnered from bio-loggers. In these models, haul-out
probabilities were expressed as a function of predictive covariates, such as time of day,
day-of-year, sex, age class, and environmental conditions (e.g., [@reder2003pb],
[@bengtson2004pb], [@bengtson2005pb], [@udevitz2009cjz], and [@verhoef2014sm]). However,
sample sizes have often been low and insufficient to permit strong inference about life
history and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
Cameron's [-@bengtson2004pb] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005pb] study was based on 6 telemetered ringed seals in the
Chukchi Sea. These studies were often further limited by logistical constraints on tagging
fieldwork and the attachment duration or operational life of satellite tags.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt, which limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class were determined at the time of deployment by
experts in the field. Field identification of age class can be uncertain in some
situations. The most notable is reliably discerning subadults from adults. In the case of
ribbon seals, pelage differences between subadults and adults are distinguishable, with
subadults often having less defined ribbons. Bearded seal subadults will often have a
spotted pattern in the pelage that is not seen in adults. Spotted seal pelage is
monomorphic and can not be used to discern age class. Despite these challenges, we feel
the age classifications used in this analysis are reliable. Seals determined to be less
than one year were classified as young-of-the-year. Subadults are those seals likely
greater than one year of age but not yet sexually mature. Adults are likely sexually
mature and older than approximately four years. For those tags deployed on
young-of-the-year and transmitting into the next year, the age class was advanced to
subadult on 1 March of the following year. Table \@ref(tab:tblDeploy) provides a summary
of these deployments and data received from them.

@@

```{r seal-hours}
seal_hours_tbl <- dat.sf %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(wkofyr = lubridate::epiweek(haulout_dt),
                dayofwk = lubridate::wday(haulout_dt, label = TRUE, abbr = TRUE) %>% 
                  forcats::fct_rev()
  ) %>% 
  dplyr::group_by(species, wkofyr, dayofwk) %>% 
  dplyr::count()
```

Tags that fall off due to molt or mortality and remain on ice or land can still send data
to satellites, and we did not want to include data from these tags in our analyses. As
such, start and end times of each deployment were identified by experts who examined
several simultaneous data streams (e.g., tag locations and dive behavior) to determine
when tags ceased providing data consistent with seal behavior. Sensor readings outside of
these start and end times were discarded prior to analysis. After approximately 9 months,
7 devices deployed on the rear flipper of bearded seals reported implausible sensor data.
All data after the first instance of unrealistic values were censored from this analysis.
Figure \@ref(fig:dataCal) shows a calendar distribution of these data across the study
season for each species. Observations are concentrated in the months of May and June due
to the timing of deployment (April and May) and the timing of molt (May and June). During
molt, seals (and their attached bio-loggers) spend more time out of the water and more
data is transmitted. Molt timing also impacts when many deployments end as any tags
adhered to the hair will fall off.

```{r seal-hours-calendar, include=TRUE, fig.height = 4.75, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, a majority of deployments started in April and May. Data were collated across all years between 2005 and 2020."}
ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.35)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  #scale_y_discrete(labels = c("Sat","","","Wed","","","Sun"))+
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(
        axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "day of week", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Haul-out behavior data is concentrated in the months of May and June",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

Haul-out behavior data was recorded and transmitted via the Argos satellite network as
hourly percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by the
tag firmware every few seconds and a percent of the hour in a dry state was calculated
(Figure \@ref(fig:examplePlot)). Each data transmission message consisted of a complete
24-hour record. Tags were deployed and programmed in a manner to maximize data
transmission during the spring pupping and molting period, however, not all days during
the deployment were successfully transmitted. This is due to a variety of factors
including satellite coverage, tag availability (i.e. tags mounted to the rear flipper are
not available to the surface while the seal is at sea), tag performance, duty cycling, and
extra-terrestrial interference.

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, May, and June. Empty areas of the plot represent missing data that were not successfully transmitted from the tag."}
d <- dat.sf %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt))

p <- ggplot(d,aes(day,hour,fill=percent_dry))+
  geom_tile(color= "white",size=0) + 
  scale_fill_distiller(palette = "YlGnBu",
                       name = "hourly percent dry", aesthetics = "fill",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                           barheight = 0.5, title.hjust = 0.5)
    )
  # scale_fill_viridis_c(name = "hourly percent dry", aesthetics = "fill",
  #   guide = guide_colorbar(title.position = "bottom", barwidth = 15,
  #                          barheight = 0.5, title.hjust = 0.5)
  # )
p <- p + facet_grid(year~month)
#p <- p + scale_y_continuous(trans = "reverse")
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  ggtitle("Observation records are incomplete because not all data are transmitted",
          subtitle = "percent-dry data for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)")
p
```

@@

```{r wxCovMap, include=TRUE, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 05 May 2012. The NARR was used to predict weather covariates for each hour in a percent dry timeline record."}

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem counterintuitive
when modeling the haul-out behavior of seal species with such a close association to sea
ice; seals haul out in the presence of sea ice, and we could assess the local
concentration of sea ice during these events (see [@olnes2020]). However, a major focus of
this study is to develop models applicable for aerial survey correction factors and using
sea ice as a covariate would almost certainly bias haul-out predictions towards those
seals that are on or near ice and therefore more likely to haul out.  Since aerial surveys
can only detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of the
bio-logger deployments used in this study were single attachments to the rear flipper of
the seal and do not provide at-sea locations, limiting our ability to fully evaluate
habitat preferences related to sea-ice. It should be noted that the aerial survey counts
are modeled with sea-ice and other habitat parameters. So, any final abundance estimates
that combine models of survey counts and our model for proportion of seals hauled out on
ice do account for the strong association between these species and sea-ice in the Bering
and Chukchi seas. Lastly, our study was limited to the spring season when seal haul-out
tendencies are strongly influenced by pupping, nursing, breeding, and molt and these
drivers are likely more influential than specific sea-ice concentration. Crawford et al
[-@crawford2019] compared haul-out probability models for bearded seals and found those
that only included season (and not sea-ice concentration) where the most parsimonious. 
For these reasons, we have elected not to use sea ice concentration as a predictor for
haul-out probability; studies to evaluate habitat selection in relation to sea-ice
parameters and other environmental covariates are in progress.

@@

Models omitting year effects suggested that day-of-year, solar hour, age-sex class,
temperature, and wind substantially altered haul-out behavior of all three species, with
$F$ tests producing $p$-values less than 0.05 for variables embodying these effects and/or
their interactions. Haul-out probabilities typically increased throughout March and April,
reaching a peak in May and early June before declining again. Diurnal patterns were
present, with maximum haul-out behavior centered around local solar noon.

@@

Ribbon seals show a pattern of gradually increasing haul-out probability in April that
peaks in May for subadults and in June for adults. The behavior is clearly centered around
local solar noon and expands to other hours later in the season as seals enter their
molting period. Subadults showed an earlier start and more intense haul-out activity in
April and May. The young-of-the-year records begin after weaning and the model predictions
demonstrate the ontogeny of in-water activities (e.g. diving, foraging) in May. Adult
females have a more protracted season compared to males and more time spent resting in
June and July. The model suggests adult male ribbon seals complete their molt by the end
of June. Lastly, for adults of both sexes there is some indication for a shift to a
nocturnal haul-out pattern in late June and July.

```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOCal", fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model."}
plot_df <- ribbon_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label = TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug")

p <- ggplot(plot_df, aes(day,solar_hour,fill = ho_prob)) +
  geom_tile(color = "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour = "white")) +
  theme(axis.ticks = element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Ribbon seals rest on sea ice in late spring and around solar noon.")
p
```

```{r}
ribbon_type3 <- tibble(
  fit_ribbon$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>% 
  column_to_rownames(var = "Effect") 
ribbon_type3 <- setNames(split(ribbon_type3, 
                               seq(nrow(ribbon_type3))), 
                         rownames(ribbon_type3))

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season that is
less intensely centered on solar noon. Adults of both sexes spend considerable time in
April and May resting on the ice. As with ribbon seals, the young-of-the-year records
begin after weaning and the model predictions demonstrate the ontogeny of in-water
activities (e.g. diving, foraging) in May.

>>>>

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring
(April-June) as sea ice begins to melt and retreat northward. Rapid acceleration of
climate change in Arctic ecosystems is therefore of concern as the quantity and quality of
suitable habitat is forecast to decrease. In this paper, we use data collected from
satellite-linked bio-loggers deployed between 2005 and 2020 to investigate the seasonal
timing and environmental factors affecting sea ice use by seals (specifically, bearded,
ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to providing
baseline data on phenology, these data also allow us to quantify 'availability', which is
needed to accurately estimate abundance from aerial survey counts of seals basking on ice
(i.e., to correct for the proportion of animals that are in the water while surveys are
conducted). Using generalized linear mixed pseudo-models to properly account for temporal
autocorrelation, we fit models with a variety of covariates (e.g., day-of-year, solar
hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to
examine their ability to explain variation in hourly haul-out records. We found evidence
for strong diurnal and within-season patterns in haul-out behavior, as well as strong
weather effects (particularly wind and temperature). In general, seals were more likely to
rest on ice in the middle of the day and when wind speed was low and temperatures were
higher. Haul-out propensity increased through March and April, peaking in May and early
June before declining again. The timing and frequency of haul-out events also varied based
on species and age-sex class. For ribbon and spotted seals, models with year effects were
highly supported, indicating that the timing and magnitude of haul-out behavior varied
among years. However, we did not find evidence that haul-out timing was linked to annual
sea ice extent. Our analysis emphasizes the importance of accounting for variation in
haul-out behavior when interpreting the number of seals counted in aerial surveys.

@@

The global climate disruption is causing considerable reduction in Arctic sea ice extent,
volume, and seasonal presence [@kwok2018erl, @meier2014rg, @wang2017dsrpitsio]. These
changes have ripple effects on Arctic organisms, ecosystems, and the human communities who
live in the region [@huntington2020ncc]. Such disruptions are a particular cause of
concern for ice-associated seals (ringed (*Pusa hispida*), bearded (*Erignathus
barbatus*), spotted (*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that
depend on spring and early summer sea ice (March-June) in the Bering and Chukchi seas as a
platform for important functions, such as pupping, nursing, breeding, and molting
[@boveng2009a; @boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge
gaps complicate predictions about the ultimate effects of changes in sea ice on the
behavior, abundance, and distribution of these seals. Knowledge about evolutionary
constraints on the timing of reproductive and molting behavior is generally lacking, so it
is difficult to predict the readiness with which ice-associated seal species will be able
to adapt to future changes (e.g., by adjusting pupping or molting schedules to earlier
dates or different locales). Additionally, trends in abundance of these species are
largely unknown, so it is difficult to say what effect, if any, declines in sea ice
habitat have had, or will have, on seal densities.

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals rest
on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The remoteness of the Bering and Chukchi
seas mean direct scientific observation of seal behavior is impractical. Bio-logging
devices are, thus, especially useful tools for collecting key information on movement and
haul-out behavior for these species.

Bio-logging records of time spent out of the water provide valuable data for identifying
covariates that explain variation in haul-out behavior. For instance, in the Antarctic,
Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to demonstrate greater
haul-out propensity in juvenile crabeater seals (*Lobodon carcinophaga*) than adults, with
highest probabilities in February and at times close to solar noon. In the Arctic, Von
Duyke et al. [-@vonduyke2020ee] used seal-borne, satellite-linked loggers to corroborate
seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly et al. [-@kelly2010] using labor-intensive VHF radio tags
and direct observation. Bengtson et al. [-@bengtson2005pb] documented a higher propensity
for ringed seal basking near solar noon, as did Ver Hoef et al. [-@verhoef2014sm] in an
analysis of bearded, ribbon, and spotted seals using much larger sample sizes. Olnes et
al. [-@olnes2020pb] showed that the proportion of time bearded seals spent hauled out
progressively increased through spring and summer. And, Ver Hoef et al. [-@verhoef2014sm]
found haul-out probabilities increased gradually starting in March and peaking in May and
June for bearded, ribbon, and spotted seals.

While bio-logging of haul-out patterns clearly has utility for description of
ice-associated seals' natural history and ecology, some studies have focused, as we do
here, specifically on quantifying the probability of seals being hauled out as a means of
accounting for seals in the water, and therefore missed, during aerial surveys. These
studies typically use logistic regression-style analyses to estimate the time-specific
probability of being hauled out base on 'wet/dry' data relayed by bio-loggers. In these
models, haul-out probabilities were expressed as a function of predictive covariates, such
as time of day, day-of-year, sex, age class, and environmental conditions (e.g.,
[@reder2003pb], [@bengtson2004pb], [@bengtson2005pb], [@udevitz2009cjz], [@verhoef2014sm],
and [@southwell2008]). However, sample sizes have often been low and insufficient to
permit strong inference about life history and/or seasonal variation in haul-out
probabilities. For instance, Bengtson and Cameron's [-@bengtson2004pb] study included 5
adult and 2 juvenile crabeater seals, while Bengtson et al.'s [-@bengtson2005pb] study was
based on 6 telemetered ringed seals in the Chukchi Sea. These studies were often further
limited by logistical constraints on tagging fieldwork and the attachment duration or
operational life of satellite tags. In this study, we address some of these limitations by
deploying small bio-loggers designed for attachment on a portion of the study individuals.
These devices are designed to collect data through the molt period (when those adhered to
the hair would fall off) and, in some situations, provide multiple years of data.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was filtered to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt, which limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class (non-dependent *young-of-the-year*, sexually
immature *subadults*, and mature *adults*) were determined at the time of deployment by
subjective judgement based on length, and pelage characteristics for some species. Field
identification of age class can be inexact, particularly when discerning subadults from
adults. In the case of ribbon seals, pelage differences between subadults and adults are
distinguishable, with subadults often having less defined ribbons. Bearded seal subadults
will often have a spotted pattern in the pelage that is not seen in adults. Spotted seal
pelage is monomorphic and can not be used to discern age class. Despite these challenges,
we feel the age classifications used in this analysis are useful categories in testing for
age-related effects on haul-out behavior. Seals determined to be less than one year were
classified as young-of-the-year. Subadults are those seals likely greater than one year of
age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on 1 March of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

```{r seal-hours}
seal_hours_tbl <- dat.sf %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(wkofyr = lubridate::epiweek(haulout_dt),
                dayofwk = lubridate::wday(haulout_dt, label = TRUE, abbr = TRUE) %>% 
                  forcats::fct_relevel("Sun","Mon","Tue","Wed","Thu","Fri","Sat")
  ) %>% 
  dplyr::group_by(species, wkofyr, dayofwk) %>% 
  dplyr::count()
```

Tags that fall off due to molt or mortality and remain on ice or land can still send data
to satellites, and we did not want to include data from these tags in our analyses. As
such, end times of each deployment were identified by experts who examined several
simultaneous data streams (e.g., tag locations and dive behavior) to determine when tags
ceased providing data consistent with seal behavior. Sensor readings outside of these
start and end times were discarded prior to analysis. After approximately 9 months, 7
devices deployed on the rear flipper of bearded seals reported implausible sensor data.
All data after the first instance of unrealistic values were censored from this analysis.
Figure \@ref(fig:dataCal) shows a calendar distribution of these data across the study
season for each species. Observations are concentrated in the months of May and June due
to the timing of deployment (April and May) and the timing of molt (May and June). During
molt, seals (and their attached bio-loggers) spend more time out of the water and more
data is transmitted. Molt timing also impacts when many deployments end as any tags
adhered to the hair will fall off.

```{r seal-hours-calendar, include=TRUE, fig.height = 4.75, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, a majority of deployments started in April and May. Data were collated across all years between 2005 and 2020."}
ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.3), breaks = c(100,500,1000,1500)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  scale_y_discrete(limits = rev) +
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(
        #axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Haul-out behavior data is concentrated in the months of May and June",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

Haul-out behavior data was recorded and transmitted via the Argos satellite network as
hourly percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by the
tag firmware every few seconds and a percent of the hour in a dry state was calculated
(Figure \@ref(fig:examplePlot)). Each data transmission message consisted of a complete
24-hour record. Tags were, generally, deployed and programmed in a manner to maximize data
transmission during the spring pupping and molting period, however, not all days during
the deployment were successfully transmitted. This is due to a variety of factors
including satellite coverage, tag availability (i.e. tags mounted to the rear flipper are
not available to the surface while the seal is at sea), tag performance, duty cycling, and
extra-terrestrial interference.

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, and May. Empty areas of the plot within a month represent missing data that were not successfully transmitted from the tag."}
d <- dat.sf %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt)) %>% 
  filter(month != 'Jun')

p <- ggplot(d,aes(day,hour,fill=percent_dry))+
  geom_tile(color= "white",size=0) + 
  scale_fill_distiller(palette = "YlGnBu",
                       name = "hourly percent dry", aesthetics = "fill",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                           barheight = 0.5, title.hjust = 0.5)
    )

p <- p + facet_grid(year~month, scales = "free_x")
#p <- p + scale_y_continuous(trans = "reverse", expand = c(0,0))
p <- p + scale_x_continuous(breaks = c(5,15,25), expand = c(0,0)) +
  scale_y_continuous(breaks = c(4,12,20))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  theme(panel.spacing = unit(0,"line")) +
  ggtitle("Observation records are incomplete because not all data are transmitted",
          subtitle = "percent-dry data for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)")
p
```

@@

```{r wxCovMap, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 05 May 2012. The NARR was used to predict weather covariates for each hour in a percent dry timeline record."}

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem counterintuitive
when modeling the haul-out behavior of seal species with such a close association to sea
ice; seals haul out in the presence of sea ice, and we could assess the local
concentration of sea ice during these events (see [@olnes2020]). However, a major focus of
this study is to develop models applicable for aerial survey correction factors and using
sea ice as a covariate would almost certainly bias haul-out predictions towards those
seals that are on or near ice and therefore more likely to haul out. Since aerial surveys
can only detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of the
bio-logger deployments used in this study were single attachments to the rear flipper of
the seal and do not provide at-sea locations, limiting our ability to fully evaluate
habitat preferences related to sea-ice. It should be noted that the aerial survey counts
are modeled with sea-ice and other habitat parameters. So, any final abundance estimates
that combine models of survey counts and our model for proportion of seals hauled out on
ice do account for the strong association between these species and sea-ice in the Bering
and Chukchi seas. Lastly, our study was limited to the spring season when seal haul-out
tendencies are strongly influenced by pupping, nursing, breeding, and molt and these
drivers are likely more influential than specific sea-ice concentration. Crawford et al
[-@crawford2019] compared haul-out probability models for bearded seals and found those
that only included season (and not sea-ice concentration) were the most parsimonious. For
these reasons, we have elected not to use sea ice concentration as a predictor for
haul-out probability; studies to evaluate habitat selection in relation to sea-ice
parameters and other environmental covariates are in progress.

@@

Models omitting year effects suggested that day-of-year, solar hour, age-sex class,
temperature, and wind substantially influenced haul-out behavior of all three species,
with $F$ tests producing $p$-values less than 0.05 for variables embodying these effects
and/or their interactions. Haul-out probabilities typically increased throughout March and
April, reaching a peak in May and early June before declining again. Diurnal patterns were
present, with maximum haul-out behavior centered around local solar noon.

@@

Ribbon seals show a pattern of gradually increasing haul-out probability in April that
peaks in late May for subadults and in early June for adults (Figure
\@ref(fig:ribbonHOCal)). The behavior is clearly centered around local solar noon and
expands to other hours later in the season as seals enter their molting period. Subadults
showed an earlier start and more intense haul-out activity in April and May. The
young-of-the-year records begin after weaning and the model predictions demonstrate the
ontogeny of in-water activities (e.g. diving, foraging) in May. Adult females have a more
protracted haul-out season compared to males, and more time spent resting in June and
July. The model suggests adult male ribbon seals complete their molt by the end of June.
Lastly, for adults of both sexes there is some indication for a shift to a crepuscular
haul-out pattern in late June and July.

```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOCal", fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}
plot_df <- ribbon_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label = TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug")

p <- ggplot(plot_df, aes(day,solar_hour,fill = ho_prob)) +
  geom_tile(color = "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30)) +
  scale_y_continuous(breaks = c(4,12,20))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour = "white")) +
  theme(axis.ticks = element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Ribbon seals rest on sea ice in late spring and around solar noon.")
p
```

```{r type3}
ribbon_type3 <- tibble(
  fit_ribbon$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>% 
  column_to_rownames(var = "Effect") 
ribbon_type3 <- setNames(split(ribbon_type3, 
                               seq(nrow(ribbon_type3))), 
                         rownames(ribbon_type3))

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season that is
less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)). Adults of both sexes spend considerable time in
April and May resting on the ice. As with ribbon seals, the young-of-the-year records
begin after weaning and the model predictions demonstrate the ontogeny of in-water
activities (e.g. diving, foraging) in May.
